This code will take all the different cleaned data files and bring them together for easier EDA and future use.

```{r}
library(dplyr)
library(stringr)
library(tidyverse)
```


First we are going to do it all by year:

Read in all in all the variables from their csvs
```{r}
# read in sports expendature data - our independent variables
sports_df_18 <- read.csv('data/clean_sports_expend_data/sports_expenditures_2018.csv')
sports_df_19 <- read.csv('data/clean_sports_expend_data/sports_expenditures_2019.csv')
sports_df_20 <- read.csv('data/clean_sports_expend_data/sports_expenditures_2020.csv')
sports_df_21 <- read.csv('data/clean_sports_expend_data/sports_expenditures_2021.csv')
sports_df_22 <- read.csv('data/clean_sports_expend_data/sports_expenditures_2022.csv')

# read in our dependent variables
discounts_df <- read.csv('data/clean_fin_data/2018-2022_discounts_to_tuiton.csv')
gifts_df <- read.csv('data/clean_fin_data/2018-2022_gifts.csv')
research_df <- read.csv('data/clean_fin_data/2018-2022_research_spent.csv')
grad_df <- read.csv('data/clean_pop_and_grad_data/graduation_rates_2018-2022.csv')

# read in our population data for normalizing
pop_df <- read.csv('data/clean_pop_and_grad_data/school_pops_2018-2023.csv')
```

Function to Remove dollar sign and make it a numeric
```{r}
remove_dollar<- function(x) {  
   # Remove dollar signs and commas, and convert to numeric
  if (is.character(x)) {
    # Remove dollar sign and commas (optional)
    x <- gsub("[\\$,]", "", x)  # Remove $ and comma (if any)
    
    # Convert the cleaned string to numeric
    x <- as.numeric(x)
  }
  
  # Return the processed column
  return(x)
}
```


Sum all the independent variables to also have a total column
```{r}
# turn each into a numeric and then add a summation column
sports_df_18[, 3:27] <- lapply(sports_df_18[, 3:27], remove_dollar)
sports_df_18$total_expendature <- rowSums(sports_df_18[ , 3:27], na.rm=TRUE)

sports_df_19[, 3:27] <- lapply(sports_df_19[, 3:27], remove_dollar)
sports_df_19$total_expendature <- rowSums(sports_df_19[ , 3:27], na.rm=TRUE)

sports_df_20[, 3:27] <- lapply(sports_df_20[, 3:27], remove_dollar)
sports_df_20$total_expendature <- rowSums(sports_df_20[ , 3:27], na.rm=TRUE)

sports_df_21[, 3:27] <- lapply(sports_df_21[, 3:27], remove_dollar)
sports_df_21$total_expendature <- rowSums(sports_df_21[ , 3:27], na.rm=TRUE)

sports_df_22[, 3:27] <- lapply(sports_df_22[, 3:27], remove_dollar)
sports_df_22$total_expendature <- rowSums(sports_df_22[ , 3:27], na.rm=TRUE)
```

Merge all dependent varaibles by year
```{r}
# initialize the variables
dep_vars_2018 <- data.frame()
dep_vars_2019 <- data.frame()
dep_vars_2020 <- data.frame()
dep_vars_2021 <- data.frame()
dep_vars_2022 <- data.frame()

# the yearly data starts at columns 4
combine_columns <- function(df1, df2, df3, df4) {
# Extract columns 1 to 3 (same in all dataframes)
  col1 <- df1[[1]]

  # Combine the 4th column from all four data frames into one data frame
  df_col4 <- data.frame(col1, df1[[4]], df2[[4]], df3[[4]], df4[[4]])
  colnames(df_col4) <- c("UNITID", "discounts_to_tuition", "gifts", "reseach_expendatures", "graduation_rate")
  
  # Combine the 5th column from all four data frames into one data frame
  df_col5 <- data.frame(col1, df1[[5]], df2[[5]], df3[[5]], df4[[5]])
  colnames(df_col5) <- c("UNITID", "discounts_to_tuition", "gifts", "reseach_expendatures", "graduation_rate")
  
  # Combine the 6th column from all four data frames into one data frame
  df_col6 <- data.frame(col1, df1[[6]], df2[[6]], df3[[6]], df4[[6]])
  colnames(df_col6) <- c("UNITID", "discounts_to_tuition", "gifts", "reseach_expendatures", "graduation_rate")
  
  # Combine the 7th column from all four data frames into one data frame
  df_col7 <- data.frame(col1, df1[[7]], df2[[7]], df3[[7]], df4[[7]])
  colnames(df_col7) <- c("UNITID", "discounts_to_tuition", "gifts", "reseach_expendatures", "graduation_rate")
  
  # Combine the 8th column from all four data frames into one data frame
  df_col8 <- data.frame(col1, df1[[8]], df2[[8]], df3[[8]], df4[[8]])
  colnames(df_col8) <- c("UNITID", "discounts_to_tuition", "gifts", "reseach_expendatures", "graduation_rate")
  
  # Return the five data frames as a list
    return(list(df_col4 = df_col4, df_col5 = df_col5, df_col6 = df_col6, df_col7 = df_col7, df_col8 = df_col8))
}


# call the function with the dataframes
result <- combine_columns(discounts_df, gifts_df, research_df, grad_df)

# Accessing each column's data frame:
dep_vars_2018 <- result$df_col4
dep_vars_2019 <- result$df_col5
dep_vars_2020 <- result$df_col6
dep_vars_2021 <- result$df_col7
dep_vars_2022 <- result$df_col8
```

Merge the datasets and save to files
```{r}
all_2018 <- merge(sports_df_18, dep_vars_2018, by = "UNITID")
all_2019 <- merge(sports_df_19, dep_vars_2019, by = "UNITID")
all_2020 <- merge(sports_df_20, dep_vars_2020, by = "UNITID")
all_2021 <- merge(sports_df_21, dep_vars_2021, by = "UNITID")
all_2022 <- merge(sports_df_22, dep_vars_2022, by = "UNITID")
```

Write all the dataframes to files
```{r}
write.csv(all_2018,'data/all_variables/by_year/2018_all_vars.csv', row.names = FALSE)
write.csv(all_2019,'data/all_variables/by_year/2019_all_vars.csv', row.names = FALSE)
write.csv(all_2020,'data/all_variables/by_year/2020_all_vars.csv', row.names = FALSE)
write.csv(all_2021,'data/all_variables/by_year/2021_all_vars.csv', row.names = FALSE)
write.csv(all_2022,'data/all_variables/by_year/2022_all_vars.csv', row.names = FALSE)
```

Now combine for an all years file
```{r}
#put all data frames into list
df_list <- list(all_2018, all_2019, all_2020, all_2021, all_2022)

#merge all data frames in list
all_years <- df_list |> reduce(full_join, by='UNITID')

# and write to a csv
write.csv(all_years,'data/all_variables/all_years/2018-2022_all_vars.csv', row.names = FALSE)
```

Now to normalize all the numbers for the population(fall enrollment)
```{r}
norm_all_2018 <- all_2018
norm_all_2018[, 4:32] <-  round((all_2018[, 4:32] / pop_df$X2018_total_enroll),4)

norm_all_2019 <- all_2019
norm_all_2019[, 4:32] <- round((all_2019[, 4:32] / pop_df$X2019_total_enroll),4)

norm_all_2020 <- all_2020
norm_all_2020[, 4:32] <- round((all_2020[, 4:32] / pop_df$X2020_total_enroll),4)

norm_all_2021 <- all_2021
norm_all_2021[, 4:32] <- round((all_2021[, 4:32] / pop_df$X2021_total_enroll),4)

norm_all_2022 <- all_2022
norm_all_2022[, 4:32] <- round((all_2022[, 4:32] / pop_df$X2022_total_enroll),4)
```


Write all the dataframes to files
```{r}
write.csv(norm_all_2018,'data/all_variables/by_year/2018_all_vars_norm.csv', row.names = FALSE)
write.csv(norm_all_2019,'data/all_variables/by_year/2019_all_vars_norm.csv', row.names = FALSE)
write.csv(norm_all_2020,'data/all_variables/by_year/2020_all_vars_norm.csv', row.names = FALSE)
write.csv(norm_all_2021,'data/all_variables/by_year/2021_all_vars_norm.csv', row.names = FALSE)
write.csv(norm_all_2022,'data/all_variables/by_year/2022_all_vars_norm.csv', row.names = FALSE)
```

Now combine for an all years file
```{r}
#put all data frames into list
df_list_norm <- list(norm_all_2018, norm_all_2019, norm_all_2020, norm_all_2021, norm_all_2022)

#merge all data frames in list
all_years_norm <- df_list_norm |> reduce(full_join, by='UNITID')

# and write to a csv
write.csv(all_years_norm,'data/all_variables/all_years/2018-2022_all_vars_norm.csv', row.names = FALSE)
```