File to perform hypothesis testing for coaching and support staff compensation

1. T-Test: investigate whether institutions with higher athletic spending offer significantly higher support staff compensation compared to those with lower spending
    
Data Categorization: High Spending and Low Spending: 
^ based on median athletic spending (total_ops_2018)

Null (H0): The mean support staff compensation is the same or lower for high spending institutions compared to low spending institutions.

Alternative Hypothesis (H1): The mean support staff compensation is significantly higher for high spending institutions.

```{r}
library(tidyverse)


data <- read.csv("data/all_variables/all_years/2018-2022_all_vars_norm.csv")

# Create spending tiers (high vs. low) based on median athletic spending
median_spending <- median(data$total_ops_2018, na.rm = TRUE)
data <- data %>%
  mutate(spending_tier = ifelse(total_ops_2018 >= median_spending, "High", "Low"))

# Perform Welch's T-Test
t_test_result <- t.test(support_comp_2018 ~ spending_tier, data = data, 
                        alternative = "greater", var.equal = FALSE)

# Display the test results
print(t_test_result)


```

Result: 

Test Statistic (t): 8.3378

The larger value of 8.3378 indicates a stronger deviation from the null hypothesis.

p-value = 1.113e-11:

Since p < 0.05, we conlude that institutions with high spending have significantly higher support staff compensation.

95 percent confidence interval:
 54.24551      Inf

Mean in group High  mean in group Low
          93.18476           25.32612

Difference in means: $67,858 shows that high-spending institutions allocate significantly more to support staff compensation.

2. Chi-Square: Evaluate whether spending tiers (low, medium, high) are associated with support staff compensation levels (below or above $50k)

Null (H0): Spending tiers(low,medium, high) are independent of whether support staff compensation exceeds a specific threshold ($50k)
Alternative Hypothesis (H1): Spending tiers are associated with support staff compensation levels.

Categorization:

Spending Tiers: Divide institutions into low, medium, and high spending tiers based on quantiles of total_ops_2018.

Compensation Levels: Categorize support staff compensation (support_comp_2018) as "Above" or "Below" a threshold ($50k)

```{r}

library(tidyverse)

data <- read.csv("data/all_variables/all_years/2018-2022_all_vars_norm.csv")

# Create spending tiers based on quantiles of total_ops_2018
data <- data %>%
  mutate(spending_category = cut(total_ops_2018,
                                 breaks = quantile(total_ops_2018, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE),
                                 labels = c("Low", "Medium", "High")))

# Categorize support staff compensation as Above or Below $50,000
data <- data %>%
  mutate(comp_above_50k = ifelse(support_comp_2018 > 50, "Above", "Below"))

# Create a contingency table
contingency_table <- table(data$spending_category, data$comp_above_50k)

# Display the contingency table
print(contingency_table)

# Perform the Chi-Square Test
chi_square_result <- chisq.test(contingency_table)

# Display the results of the test
print(chi_square_result)

```
Contingency Table

Above Below
  Low        0    33
  Medium    12    24
  High      34     0

Observations: All 33 instituions fall into the below $50,000 category

Test Statistic: X-squared = 70.631

Provides that spending tiers (low, medium, high) are not independent of support staff compensation levels (above/below $50k)

^ Quantifying a deviation from the null hypothesis, confirming that higher spending is strongly associated with higher compensation.

p-value = 4.6e-16

P-value is smaller than the standard significance threshold (0.05), providing strong evidence against the null.

Conclusion: There is a statistically significant association between spending tiers and support staff compensation levels. (reject null)


3. Confidence Intervals for Mean Coach Compensation:

Purpose: Estimate the mean head coach compensation for high and low spending institutions with a 95% confidence interval (CI).

Allows for comparison between high and low spending teirs to observe disparities in head coach compensation.

Null(H0): The true mean head coach compensation for high-spending institutions is equal to that of low-spending institutions.

Alternative Hypothesis(H1): The true mean head coach compensation for high spending institutions is significantly different from that of low-spending instituions.


```{r}

data <- read.csv("data/all_variables/all_years/2018-2022_all_vars_norm.csv")

# Calculate the median of total_ops_2018
median_spending <- median(data$total_ops_2018, na.rm = TRUE)

# Subset data directly for high and low spending institutions
high_spending_coach <- data$coach_comp_2018[data$total_ops_2018 >= median_spending]
low_spending_coach <- data$coach_comp_2018[data$total_ops_2018 < median_spending]

# Perform t-tests to calculate 95% confidence intervals
ci_high <- t.test(high_spending_coach, conf.level = 0.95)$conf.int
ci_low <- t.test(low_spending_coach, conf.level = 0.95)$conf.int

# Display results
cat("95% CI for High Spending Institutions:", ci_high, "\n")
cat("95% CI for Low Spending Institutions:", ci_low, "\n")



```

Results:

95% CI for High Spending Institutions: 83.89135, 114.3667

The true mean head coach compensation for instituions in the high spending tier is expected to fall between approximately $83,891 and $114,367 with 95% confidence.

95% CI for Low Spending Institutions: 27.10594, 34.26166

The true mean head coach compensation for institutions in the low spending tier is expected to fall between approximately $27,106 and $34,262 with 95% confidence.

4. Permutation Test for Coach Compensation Across Spending Tiers
   
Purpose: Test whether the observed difference in mean head coach compensation between high and low spending institutions is statistically significant using a non-parametric approach that does NOT rely on data distribution assumptions.

Null(H0): The mean head coach compensation for high-spending institutions is equal to the mean head coach compensation for low-spending institutions. Any observed difference in means is due to random chance.

Alternative(H1): The mean head coach compensation for high-spending institutions is significantly different from the mean head coach compensation for low-spending institutions. This difference is not due to random chance.

```{r}
set.seed(123)  # For reproducibility


data <- read.csv("data/all_variables/all_years/2018-2022_all_vars_norm.csv")

# Calculate the median spending to split into high and low groups
median_spending <- median(data$total_ops_2018, na.rm = TRUE)

# Subset the data based on spending levels
high_spending_coach <- data$coach_comp_2018[data$total_ops_2018 >= median_spending]
low_spending_coach <- data$coach_comp_2018[data$total_ops_2018 < median_spending]

# Remove missing values if any
high_spending_coach <- na.omit(high_spending_coach)
low_spending_coach <- na.omit(low_spending_coach)

# Calculate the observed difference in means
obs_diff <- mean(high_spending_coach) - mean(low_spending_coach)

# Combine all compensation values
all_coach_comp <- c(high_spending_coach, low_spending_coach)
n_high <- length(high_spending_coach)

# Perform permutations
perm_diffs <- replicate(10000, {
  permuted <- sample(all_coach_comp)  # Shuffle the data
  high_perm <- permuted[1:n_high]    # First n_high values as high group
  low_perm <- permuted[(n_high + 1):length(permuted)]  # Remaining as low group
  mean(high_perm) - mean(low_perm)  # Calculate mean difference
})

# Calculate the p-value
p_value <- mean(perm_diffs >= obs_diff)

# Display the results
cat("Observed Difference:", obs_diff, "\n")
cat("P-value from Permutation Test:", p_value, "\n")


```

Observed Difference: 68.44523

Value represents diffference in mean head coach compensation between high-spending and low-spending instituions. 

^ Institutions in the high spending teir pay $68,445 more to their head coaches compared to those in the low spending tier.

P-value from Permutation Test: 0

Null (H0): There is no significant difference in mean head coach compensation between high-spending and low-spending institutions.
Alternative Hypothesis (H1): High-spending institutions have a significantly higher mean head coach compensation than low-spending institutions.

5. T-Test

Compares the mean compensation for support staff (support_comp_2018) and head coaches (coach_comp_2018). The t-test is appropriate because the two groups are distinct (support staff and coaches) and compensation data is independent.

Purpose: to assess whether there is a statistically significant difference between the mean compensation of support staff and head coaches.

Null Hypothesis (H0): The mean compensation for support staff is equal to the mean compensation for head coaches.
Alternative Hypothesis (H1): The mean compensation for support staff is significantly different from the mean compensation for head coaches.

```{r}

library(tidyverse)

data <- read.csv("data/all_variables/all_years/2018-2022_all_vars_norm.csv")

# Remove missing values in compensation columns
data <- data %>%
  filter(!is.na(support_comp_2018) & !is.na(coach_comp_2018))

# Perform the independent t-test
t_test_result <- t.test(data$support_comp_2018, data$coach_comp_2018, 
                        alternative = "two.sided",  # Two-tailed test
                        var.equal = FALSE)          # Welch's t-test for unequal variances

# Display the test result
print(t_test_result)

```

Results:

t = -0.77573

A negative value indicates the mean compensation for support staff (x) is slightly lower than for head coaches (y), but the difference is not large enough to be significant.

p-value = 0.4388

Since p > 0.05, we say there is no statistically significant difference between the mean compensation for support staff and head coaches. (reject null)

95% Confidence Interval: [-20.078796, 8.739285]

The interval contains 0, further supporting the conclusion that there is no statistically significant difference.

sample estimates:
mean of x mean of y 
 58.77004  64.43980

The average compensation for support staff is $58,770 and $64,440 for head coaches. 


