---
title: "proj_eda Normalized Data"
author: Group 7

Output: html
format:
  html:
    embed-resources: true
    df-print: kable
---

```{r, echo=FALSE, message=FALSE}
library(tidyverse) |> suppressPackageStartupMessages()
library(ggplot2) |> suppressPackageStartupMessages()
library(dplyr) |> suppressPackageStartupMessages()
library(reshape2) |> suppressPackageStartupMessages()
```

## Perform EDA (Exploratory Data Analysis) using some Data Visualizations.

```{r}
merged_df <- read.csv("./all_years/all_yrs.csv")
norm_df <- read.csv("./all_years/all_yrs_norm.csv")
```

```{r, echo=FALSE, message=FALSE}
sample_n(norm_df, 5)
```

```{r}
# Gather data into long format for easier plotting
long_data <- norm_df %>%
  select(UNITID, total_ops_2018, total_ops_2019, total_ops_2020, total_ops_2021, total_ops_2022) %>%
  pivot_longer(cols = starts_with("total_ops"), 
               names_to = "year", 
               values_to = "total_ops") %>%
  mutate(year = gsub("total_ops_", "", year)) 

# Summarize total athletic spending by year
summary_data <- long_data %>%
  group_by(year) %>%
  summarize(mean_total_ops = mean(total_ops, na.rm = TRUE))

# Create barplot
ggplot(data = summary_data, aes(x = year, y = mean_total_ops, fill = year)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "black") +
  labs(title = "Average Total Athletic Spending Across Years (2018–2022)",
       x = "Year",
       y = "Mean Total Athletic Spending",
       caption = "Data: Normalized for Student Body Population",
       fill = "Year") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") 
```

Make a Means dataframe for means across 5 years of the variables
```{r}
# Create a dataframe for calculating means for the specified columns
means_df <- norm_df %>%
  rowwise() %>%
  mutate(
    coach_comp_third = mean(c(coach_comp_third_2018, coach_comp_third_2019, coach_comp_third_2020, coach_comp_third_2021, coach_comp_third_2022), na.rm = TRUE),
    support_comp_third = mean(c(support_comp_third_2018, support_comp_third_2019, support_comp_third_2020, support_comp_third_2021, support_comp_third_2022), na.rm = TRUE),
    coach_comp = mean(c(coach_comp_2018, coach_comp_2019, coach_comp_2020, coach_comp_2021, coach_comp_2022), na.rm = TRUE),
    guarantees = mean(c(guarantees_2018, guarantees_2019, guarantees_2020, guarantees_2021, guarantees_2022), na.rm = TRUE),
    total_ops = mean(c(total_ops_2018, total_ops_2019, total_ops_2020, total_ops_2021, total_ops_2022), na.rm = TRUE),
    aid = mean(c(aid_2018, aid_2019, aid_2020, aid_2021, aid_2022), na.rm = TRUE),
    support_comp = mean(c(support_comp_2018, support_comp_2019, support_comp_2020, support_comp_2021, support_comp_2022), na.rm = TRUE),
    severance = mean(c(severance_2018, severance_2019, severance_2020, severance_2021, severance_2022), na.rm = TRUE),
    recruiting = mean(c(recruiting_2018, recruiting_2019, recruiting_2020, recruiting_2021, recruiting_2022), na.rm = TRUE),
    team_travel = mean(c(teamtravel_2018, teamtravel_2019, teamtravel_2020, teamtravel_2021, teamtravel_2022), na.rm = TRUE),
    equip = mean(c(equip_2018, equip_2019, equip_2020, equip_2021, equip_2022), na.rm = TRUE),
    game_expenses = mean(c(game_expenses_2018, game_expenses_2019, game_expenses_2020, game_expenses_2021, game_expenses_2022), na.rm = TRUE),
    marketing = mean(c(marketing_2018, marketing_2019, marketing_2020, marketing_2021, marketing_2022), na.rm = TRUE),
    camps = mean(c(camps_2018, camps_2019, camps_2020, camps_2021, camps_2022), na.rm = TRUE),
    spirits = mean(c(spirit_2018, spirit_2019, spirit_2020, spirit_2021, spirit_2022), na.rm = TRUE),
    facilities = mean(c(facilities_rentals_2018, facilities_rentals_2019, facilities_rentals_2020, facilities_rentals_2021, facilities_rentals_2022), na.rm = TRUE),
    overhead = mean(c(overhead_2018, overhead_2019, overhead_2020, overhead_2021, overhead_2022), na.rm = TRUE),
    indirect_institutional = mean(c(indirect_institutional_2018, indirect_institutional_2019, indirect_institutional_2020, indirect_institutional_2021, indirect_institutional_2022), na.rm = TRUE),
    medical = mean(c(medical_2018, medical_2019, medical_2020, medical_2021, medical_2022), na.rm = TRUE),
    membership = mean(c(memberships_2018, memberships_2019, memberships_2020, memberships_2021, memberships_2022), na.rm = TRUE),
    meals = mean(c(meals_2018, meals_2019, meals_2020, meals_2021, meals_2022), na.rm = TRUE),
    otherops = mean(c(otherops_2018, otherops_2019, otherops_2020, otherops_2021, otherops_2022), na.rm = TRUE),
    bowl_expenses = mean(c(bowl_expense_2018, bowl_expense_2019, bowl_expense_2020, bowl_expense_2021, bowl_expense_2022), na.rm = TRUE),
    prof_def = mean(c(prof_def_2018, prof_def_2019, prof_def_2020, prof_def_2021, prof_def_2022), na.rm = TRUE),
    bowl_coachcomp = mean(c(bowl_coachcomp_2018, bowl_coachcomp_2019, bowl_coachcomp_2020, bowl_coachcomp_2021, bowl_coachcomp_2022), na.rm = TRUE),
    total_expenditure = mean(c(total_expendature.x, total_expendature.y, total_expendature.x.x, total_expendature.y.y), na.rm = TRUE),
    discounts_to_tuition = mean(c(discounts_to_tuition.x, discounts_to_tuition.y, discounts_to_tuition.x.x, discounts_to_tuition.y.y), na.rm = TRUE),
    gifts = mean(c(gifts.x, gifts.y, gifts.x.x, gifts.y.y), na.rm = TRUE),
    research_expenditure = mean(c(reseach_expendatures.x, reseach_expendatures.y, reseach_expendatures.x.x, reseach_expendatures.y.y), na.rm = TRUE),
    graduation_rate = mean(c(graduation_rate.x, graduation_rate.y, graduation_rate.x.x, graduation_rate.y.y), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(UNITID, school_name.x, coach_comp_third, support_comp_third, coach_comp, guarantees, total_ops, aid, 
         support_comp, severance, recruiting, team_travel, equip, game_expenses, marketing, camps, spirits, facilities,
         overhead, indirect_institutional, medical, membership, meals, otherops, bowl_expenses, prof_def, bowl_coachcomp,
         total_expenditure, discounts_to_tuition, gifts, research_expenditure, graduation_rate)
```

```{r}
sample_n(means_df, 5)
```

```{r}
# Select top 20 schools by total expenditure
top_schools <- means_df %>%
  arrange(desc(total_expenditure)) %>%
  slice_head(n = 20)
# Ensure the school names are ranked by total expenditure
top_schools <- top_schools %>%
  mutate(school_name.x = factor(school_name.x, levels = school_name.x[order(desc(total_expenditure))]))

# Barplot for top 20 schools expenditures
ggplot(top_schools, aes(x = reorder(school_name.x, total_expenditure), y = total_expenditure)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  coord_flip() +  
  labs(title = "Total Expenditure top 20 Schools",
       x = "School Name",
       y = "Total Expenditure",
       caption = "Data: Means across 2018-2022 - Normalized") 

```

```{r}
# Plot for discounts_to_tuition
plot_discounts <- ggplot(top_schools, aes(x = school_name.x, y = discounts_to_tuition)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Discounts to Tuition for Top 20 Schools",
       x = "School Name",
       y = "Discounts to Tuition",
       caption = "Data: Means across 2018-2022 - Normalized")
plot_discounts
```


```{r}
# Plot for gifts
plot_gifts <- ggplot(top_schools, aes(x = school_name.x, y = gifts)) +
  geom_bar(stat = "identity", fill = "orange", alpha = 0.8) +
  coord_flip() +
  labs(title = "Gifts for Top 20 Schools",
       x = "School Name",
       y = "Gifts",
       caption = "Data: Means across 2018-2022 - Normalized")
plot_gifts
```

```{r}
# Plot for research_expenditure
plot_research <- ggplot(top_schools, aes(x = school_name.x, y = research_expenditure)) +
  geom_bar(stat = "identity", fill = "darkgreen", alpha = 0.8) +
  coord_flip() +
  labs(title = "Research Expenditure for Top 20 Schools",
       x = "School Name",
       y = "Research Expenditure",
       caption = "Data: Means across 2018-2022 - Normalized")
plot_research
```

```{r}
# Plot for graduation_rate
plot_graduation <- ggplot(top_schools, aes(x = school_name.x, y = graduation_rate)) +
  geom_bar(stat = "identity", fill = "purple", alpha = 0.8) +
  coord_flip() +
  labs(title = "Graduation Rate for Top 20 Schools",
       x = "School Name",
       y = "Graduation Rate",
       caption = "Data: Means across 2018-2022 - Normalized") 
plot_graduation
```

```{r}
# Scatterplot: Total Operations vs Graduation Rate
ggplot(means_df, aes(x = total_ops, y = graduation_rate)) +
  geom_point(color = "darkred", alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Add a linear regression line
  labs(title = "Athletic Spending vs Graduation Rate",
       x = "Total Operations (Athletic Spending)",
       y = "Graduation Rate",
       caption = "Data: Means across 2018-2022 - Normalized") +
  theme_minimal(base_size = 12)
```

```{r}
# Select relevant numeric columns
correlation_data <- means_df %>%
  select(total_ops, gifts, research_expenditure, graduation_rate, discounts_to_tuition)

# Compute correlation matrix
correlation_matrix <- round(cor(correlation_data, use = "complete.obs"), 2)

# Convert to long format for heatmap
correlation_long <- melt(correlation_matrix)

# Heatmap with enhanced readability
ggplot(correlation_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 4) +  # Add correlation values
  scale_fill_gradient2(low = "red", high = "blue", mid = "white", midpoint = 0,
                       limits = c(-1, 1), name = "Correlation") +
  labs(title = "Correlation Between Total Sports Expendature Spending and Benefits",
       x = "",
       y = "",
       fill = "Correlation") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
        axis.text.y = element_text(size = 10))
```

## EDA Findings Summary

- Strong Positive Correlation Between total_ops and gifts (0.76):

Schools with higher athletic operational spending (total_ops) tend to receive more gifts or donations.The strong correlation between total_ops and gifts highlights athletics' role in attracting donations. Schools could leverage this by increasing visibility or engagement through athletics to boost fundraising efforts.

- Moderate Positive Correlation Between total_ops and research_expenditure (0.78):

Higher athletic spending is associated with greater research expenditures. This could indicate a general trend where schools that spend more on athletics also allocate more resources toward research, reflecting larger institutional budgets.

- Moderate Positive Correlation Between total_ops and discounts_to_tuition (0.78):

Schools with higher athletic operational spending also tend to provide more tuition discounts or scholarships.
This might reflect institutions using athletics-generated revenues to support student financial aid or related benefits. The moderate correlation between total_ops and discounts_to_tuition suggests that athletics may indirectly contribute to student financial aid, either through revenue generation or donor funding.


Low to Negative Correlation Between total_ops and graduation_rate (0.3):

- This indicates that athletic spending are not directly linked to better graduation rates. Graduation rates are likely influenced by other factors unrelated to athletics.


```{r}
# Since total_ops (athletics spending) has a strong positive correlation with gifts, a scatterplot with a regression line can show how athletics spending relates to donations.
ggplot(means_df, aes(x = total_ops, y = gifts)) +
  geom_point(color = "darkblue", alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Relationship Between Total Ops and Gifts",
       x = "Total Ops (Athletic Spending)",
       y = "Gifts (Donations)",
       caption = "Data: Means across 2018-2022 - Normalized") +
  theme_minimal(base_size = 12)
```


```{r}
# The strong correlation between gifts and research_expenditure suggests a scatterplot might reveal patterns in donor contributions and research funding.
ggplot(means_df, aes(x = gifts, y = research_expenditure)) +
  geom_point(color = "darkgreen", alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Relationship Between Gifts and Research Expenditure",
       x = "Gifts (Donations)",
       y = "Research Expenditure",
       caption = "Data: Means across 2018-2022 - Normalized") +
  theme_minimal(base_size = 12)

```


```{r}
ggplot(means_df, aes(x = total_ops, y = graduation_rate)) +
  geom_point(color = "darkred", alpha = 0.6, size = 3) +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Relationship Between Total Ops and Graduation Rate",
       x = "Total Ops (Athletic Spending)",
       y = "Graduation Rate",
       caption = "Data: Means across 2018-2022 - Normalized") +
  theme_minimal(base_size = 12)
```

